**The Role of Interoception and Theory of Mind in Deception**

### 2022-08-31 paper/manuscript_v0.Rmd
update

Interoceptive awareness (IA) is the ability to monitor, interpret, integrate and regulate the sensations and state of one's internal organs, and is closely tied to emotional and cognitive processes [@garfinkel2015knowing, \@wen]. Greater is IA associated with greater self-regulation [Weiss et al,], especially mindfulness practitioners who tend to have greater interoceptive awareness and greater ability to regulate stress and associated physiological indicators [@ardi2021effects; @mcmahon2021mindful].

@@

Since fluctuations in key physiological indicators as a means to determine whether the individual is lying is a key principle behind the polygraph test and other lie detection techniques, this paucity strengthens the need for more research examining the relationship between interoception and deception, and theory of mind, and the role it plays in moral behavior and deception. However, while there is a clear link established between interoception and deception playing a role in lie detection, limited research has been conducted examining this phenomena on its head. The act of lying is often accompanied by physiological responses such as increased heart rate, increased respiration and sweating. This research attempts to isolate the effects of theory of mind and interoception, and examine how they interact with deception. Of particular interest is how liars might utilize their interoceptive ability to self regulate in towards a more successful deception, and how this ability may compare or interact with other mechanisms at play such as theory of mind. We hypothesize that participants with greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal during the polygraph prediction. (??)

@@

##

Interoceptive awareness (IA) is the ability to monitor, interpret, integrate and regulate the sensations and state of one's internal organs, and is closely tied to emotional and cognitive processes [@garfinkel2015knowing, @chen2021emerging]. Greater IA is associated with greater self-regulation [@weiss2014interaction], especially in mindfulness practitioners who tend to have greater interoceptive awareness and greater ability to regulate stress and associated physiological indicators [@ardi2021effects; @mcmahon2021mindful].

Some research has examined how higher interoceptive accuracy, measured through heart beat detection, has been associated with vasoconstriction when faced with a liar, indicating that IA may aid in discriminating high stakes emotional lies and truths [@gunderson2021body]. It is clear that interoception may indicate on an implicit level that one is being lied to, however, there is limited research investigating how interoception may work the other way around, in favour of a liar in regulating their own physiological state that may betray their deception. The findings by @vabba2022interoceptive indicate that high cardiac interoceptive participants were more likely to be dishonest and to tell higher egoistic lies. Meanwhile, @megias2019emotionally used EEG and heart rate to find that interoceptive awareness facilitated regulation of arousal during cognitive reappraisal. There is limited research examining whether individuals with high interoceptive accuracy may be more confident in managing physiological tells associated with emotions like guilt or anxiety that the lie may be discovered through suppression or reappraisal [@fustos2013embodiment, @megias2019emotionally].

Since fluctuations in key physiological indicators as a means to determine whether the individual is lying is a key principle behind the polygraph test and other lie detection techniques, this paucity strengthens the need for more research examining the relationship between interoception and deception, and theory of mind, and the role it plays in moral behavior and deception. However, while there is a clear link established between interoception and deception playing a role in lie detection, limited research has been conducted examining this phenomena on its head. The act of lying is often accompanied by physiological responses such as increased heart rate, increased respiration and sweating. This research attempts to isolate the effects of theory of mind and interoception, and examine how they interact with deception. Of particular interest is how liars might utilize their interoceptive ability to self regulate in towards a more successful deception, and how this ability may compare or interact with other mechanisms at play such as theory of mind.

We hypothesize that participantswith greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal during the polygraph prediction. (??)

>>>>

Interoceptive awareness (IA) is the ability to monitor, interpret, integrate and regulate the sensations and state of one's internal organs, and is closely tied to emotional and cognitive processes [@garfinkel2015knowing, @chen2021emerging]. Greater is IA associated with greater self-regulation [@weiss2014interaction], especially mindfulness practitioners who tend to have greater interoceptive awareness and greater ability to regulate stress and associated physiological indicators [@ardi2021effects; @mcmahon2021mindful].

@@

The act of lying is often accompanied by physiological responses such as increased heart rate, increased respiration and sweating. Since fluctuations in key physiological indicators as a means to determine whether the individual is lying is a key principle behind the polygraph test and other lie detection techniques, there is a need for more research examining the relationship between IA, theory of mind, and deception. This research attempts to isolate the effects of theory of mind and interoception, and examine how they interact with deception. Of particular interest is how liars might utilize their interoceptive ability to self regulate in towards a more successful deception, and how this ability may compare or interact with other mechanisms at play such as theory of mind. We hypothesize that participants with greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal during the polygraph prediction. (??)

@@
